Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-10-29 demystify.tex
updates

\begin{commentbox}{Vu:}
In my opinion, requiring applicants to pay the fee helps ensure their seriousness, as it filters out non-serious candidates. Most CS programs already receive many applications and would be overwhelmed if the application process were free.  Note that if you have financial difficulties, you can ask the department for a waiver, but this is typically only granted in exceptional cases.
\end{commentbox}

@@

Most well-known and well-respected profs would \emph{not} ask you to write your own letter (in fact, even not well-known ones wouldn't do this to students they care about). This might be a common practice at specific universities and the students do not have a choice as they need the letter.  However, think about this: if a prof. does this often, then they either don't know how to write a LOR (more common than you would think) or simply do not know or care enough about you.  In any case, such LoRs are not useful and might even hurt your application.  So if you are in this situation, you should find someone else to write for you.

@@

\subsection{Help your LOR Writer}]\label{sec:help-your-LOR-writers}

@@

\subsection{Remind Your Writers}

@@

@@

@@

If your school is the top in your country but you think it might not be well-known internationally, you should talk about it in your SOP (and also ask your LOR writer to do that). Of course, if you're interested in working with Vietnamese, consider  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{CS programs in the US that have Vietnamese professors}. %It might also be helpful to have your CS dept to put itself on CSRankings so that others know about the dept, its faculty, and research.

@@

\paragraph{Bad Grades} 
While having good GPA might not help much (again, because of other much more important factors),
having bad GPA can be a \red{red flag}. In fact, many universities have a minimum GPA requirement (e.g., > 3.0) and will automatically reject applications with lower GPAs.  
If you have bad grades, you should explain them in your SOP or have your LOR writers explain them in their letters.

@@

\subsection{Selecting and Ranking Schools}\label{sec:selecting-ranking-schools}
\begin{center}
  \includegraphics[scale=0.6]{files/c1.png}
\end{center}

International students often put schools into two bins: (i) top schools that they dream about, and (ii) everything else.  They commonly use rankings from US News, which is not transparent and questionable\footnote{\url{https://cra.org/cra-statement-us-news-world-report-rankings-computer-science-universities/}}.  Sometimes they evaluate based on the reputation of the school's undergrad program or the reputation of the school's non-CS programs such as medical, math, or physics.
They even rank universities based on popular places they know in the US, e.g., California, Texas, and New York.

Instead of these superficial criteria, you should consider the CS program and the research interests of faculty members.
You can learn about these using resources such as \href{https://csrankings.org}{CSRankings.org}. You will be very surprised to learn that a school that you didn't know much about can have very strong research in your topic (and vice versa, a school you thought highly about might have no faculty working in the research field you're interested in). This is also a good way to learn about individual faculty, e.g., who works on what, and well-known CS conferences\footnote{In CS (and probably only in CS), conferences, not journals, are often the main venue to publish research findings (see \href{https://homes.cs.washington.edu/~mernst/advice/conferences-vs-journals.html}{why here}).}. %\autoref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

@@

\paragraph{What matters to you?} While many find CSRankings useful, it is not perfect, and in general, rankings are superficial, and you need to do more research to be informed and make better decisions. For example, you might prefer places with a large community from your country (e.g., Northern VA has a large population of Vietnamese). You might prefer places with high-tech companies (Seattle or Silicon Valley), with many outdoor activities (e.g., hiking, skiing), or with better weather (\emph{``PhD can be depressing, so would you rather be depressed in California or New York?''}).  You might need to think about the cost of living, e.g., living in California is more expensive than in Virginia, or places with high crime rates (note that while some universities might be in a high-crime city but the campus itself is safe).

@@

I read carefully \emph{strong LoRs} and skim weaker ones, noting either strong letters from well-known professors/researchers or generic content. I skim over \emph{CV} and look for publications, research experiences, and notable achievements, noting things like ``published papers in top venues'' or ``gold medal in an international competition.''
I read strong \emph{SOPs} thoroughly (and skim weak ones). I note whether the SOP is exciting, research-oriented, standing out, e.g., applicant is minority (\autoref{sec:urm}), and tailored to our program (e.g., if they are familiar with the work of some faculty or have talked to them). These notes are entered into the evaluation system.

@@

Most of what is written in this document applies to both domestic\footnote{For simplicity domestic means you did your undergrad (or MS) at a US university.} and international students.  However, there are some differences and benefits that domestic students should be aware of and can leverage to improve their chances of admission.

\paragraph{Standing out \autoref{sec:improve-your-chance}} There are \emph{few} domestic applications compared to international ones, i.e., domestic students are the minority (\autoref{sec:urm}) in the CS PhD application pool. Many US universities want to increase the number of domestic students in their programs (and as mentioned later, there are specific fellowships and funding for domestic students).
That makes you stand out and can help your case.

@@

While both MS and PhD programs are graduate degrees, they are \emph{very different} in terms of objective, admission requirements, course requirements, duration, and funding. This section provides guidance on applying to MS programs in CS.

@@

MS programs are much less competitive than PhD programs, i.e., you're likely to get in if you can afford it. Some refer to MS programs as a \emph{cash cow} for the department, as MS students are often not funded and have to pay tuition.

@@

\subsection{Funding}\label{sec:ms-funding}
Unlike PhD programs, which often have funding (\autoref{sec:funding}), MS students are typically not funded. The reason is because MS students do not focus on research and thus are not funded through RA and PhD students have priority for TA positions. 
\begin{itemize}
\item (RA) Profs. are not willing to take MS students as RAs because they are not around long enough to be productive. It can take a while for a student to get up to speed and start being productive, and by that time, they are about to graduate. Moreover, the goal of most MS students is to get a job, not to do research, so they are not as motivated to do research.

\item (TA) MS students are typically not given TA positions because PhDs are given priority. Many depts. do not have enough TAs for PhD students and so they cannot afford to give them to MS students.
\end{itemize}

@@

\item \textbf{Living Cost:} MS students are typically not funded, thus living cost can be a big hassle. You should consider the tuition and living expenses of the area. Note that sometimes living costs, e.g., renting, seems scary at first but students often find way to make it work, e.g., by sharing an apartment with other students.

>>>>

\begin{commentbox}{Vu:}
In my opinion, requiring applicants to pay the fee helps ensure their seriousness, as it filters out non-serious candidates. Most CS programs already receive many applications and would be overwhelmed if the application process were free.  Even with application fee the competition is already very tough, imagine if the application is free and the number of applications doubles or triples.\\

Note that if you have financial difficulties, you can ask the department for a waiver, but this is typically only granted in exceptional cases.
\end{commentbox}

@@

\begin{commentbox}[Vu:]
 Well-known and well-respected profs would \emph{not} ask you to write your own letter (in fact, even not well-known ones wouldn't do this to students they care about). This might be a common practice at specific universities and the students do not have a choice as they need the letter.  However, think about this: if a prof. does this often, then they either don't know how to write a LOR (more common than you would think) or simply do not know or care enough about you.  In any case, such LoRs are not useful and might even hurt your application.  So if you are in this situation, you should find someone else to write for you.
\end{commentbox}

@@

\subsection{Help your LOR Writer}\label{sec:help-your-LOR-writers}

@@

\subsection{Remind Your Writers}\label{sec:remind-writers}

@@

\subsection{Thank Your Writers}\label{sec:thank-writers}

After your application is submitted and your writers have submitted their letters (i.e., the wait begins), you should send a quick thank you note, which serves both as an acknowledgement that you know they have submitted the letters and as an appreciate for their help.  You should also let them know the outcome of your application, regardless of whether you are admitted or not.  In addition to being a common courtesy, this can also help maintain a good relationship with your writers, which can be useful in the future (e.g., if you need another letter for another round of applications or for a job reference).

@@

\subsection{How do you write LoR? Is it lots of work?}\label{sec:lor-writing}

If a student asks me to write letter for them, I will agree as I believe it is part of my job.  However, I will let them know if I won't be able to write a strong letter for them (e.g., they just take my class and I don't know them well enough or they have not done much research with me), and suggest they find someone else.  If they insist, then I will write the letter, but it will likely be a weak one.  However, I will still try to write something positive about the student, e.g., they are hardworking, or that I heard they are working on a research project, etc.  I usually ask the student to provide me with some details about themselves, e.g., their GPA and transcript,  and use that to write the letter.  Regardless, the letters will still be short and weak (\autoref{sec:generic-letters}) and usually take about 5-10 minutes to write.

For students who I know well, i.e., strong letters, then it will take a lot longer as it will be personalized.  While I have a general template of such strong letters, it still can take me an hour or more to write such a letter.  I also often ask the students to provide information (\autoref{sec:help-your-LOR-writers}) and what they think I should highlight in the letter. If they already have a draft of their SOP, they can also provide that to me so I can complement what they say with my own perspective.  I also offer to review their SOP and provide feedback.

%I do ask the students to waive their right to see the letter, as it is a standard practice and a way of keeping things straightforward and respectful in the whole recommendation game.  I also let them know that I will not show them the letter, even if they do not waive their right.  I also let them know that I will not write a letter for them if they do not waive their right, as I believe that such letters are not useful and might even hurt their application.

%\subsection{Will it annoy them if they have to submit many letters? }\label{sec:many-letters}
%Profs. are used to write letters for their students

@@

Many international students mistakenly believe that their school is well-known, but in fact, it is not (e.g., many Vietnamese students believe that ``Bach Khoa'' is well-known internationally, but it is not).  So you should explicitly mention that in your SOP and ask your LOR writers to do that as well. 
Of course, if you're interested in working with Vietnamese, consider  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{CS programs in the US that have Vietnamese professors}. %It might also be helpful to have your CS dept to put itself on CSRankings so that others know about the dept, its faculty, and research.

@@

\paragraph{Bad Grades} 
While having good GPA might not help much (again, because of other much more important factors),
having bad GPA can hurt your application.  Many universities have a minimum GPA requirement (e.g., $> 3.0$) and will automatically reject applications with lower GPAs.  
If you have bad grades, you should explain them in your SOP or better yet, have your LOR writers explain them in their letters if they know the reasons.

@@

\subsection{Selecting and Ranking Schools}\label{sec:selecting-ranking-schools}
\begin{center}
  \includegraphics[scale=0.5]{files/c1.png}
\end{center}

Many students just put schools into two bins: (i) top schools that they dream about, and (ii) everything else.  They commonly use rankings from US News, which is not transparent and questionable\footnote{\url{https://cra.org/cra-statement-us-news-world-report-rankings-computer-science-universities/}}.  Sometimes they evaluate based on the reputation of the school's undergrad program or the reputation of the school's non-CS programs such as medical, math, or physics.
Many international students rank universities based on popular places they know in the US, e.g., California, Texas, and New York.

Instead of these superficial criteria, you should specifically consider the CS program and the research interests of faculty members.
You can learn about these using resources such as \href{https://csrankings.org}{CSRankings.org}. You will be very surprised to learn that a school that you didn't know much about can have very strong research in your topic (and vice versa, a school you thought highly about might have no faculty working in the research field you're interested in). This is also a good way to learn about individual faculty, e.g., who works on what, and well-known CS conferences\footnote{In CS (and probably only in CS), conferences, not journals, are often the main venue to publish research findings (see \href{https://homes.cs.washington.edu/~mernst/advice/conferences-vs-journals.html}{why here}).}. %\autoref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

@@

\paragraph{What matters to you?} While many find CSRankings useful, it is still superficial as every other ranking.  You should consider other factors that matter to you.  For example, you might prefer places with a large community from your country (e.g., Northern VA has a large population of Vietnamese). You might prefer places with high-tech companies (Seattle or Silicon Valley), with many outdoor activities (e.g., hiking, skiing), or with better weather (\emph{``PhD can be depressing, so would you rather be depressed in California or New York?''}).  You might need to think about the cost of living, e.g., living in California is more expensive than in Virginia, or places with high crime rates (note that while some universities might be in a high-crime city but the campus itself is safe).

@@

I read carefully \emph{strong LoRs} and skim weaker ones, noting either strong letters from well-known professors/researchers or generic content. I skim over \emph{CV} and look for publications, research experiences, and notable achievements. I take notes of things like ``published papers in top venues'' or ``gold medal in an international competition''.

I skim weak \emph{SOPs} but read strong ones thoroughly. I note whether the SOP is exciting, research-oriented, standing out, e.g., applicant is minority (\autoref{sec:urm}), and tailored to our program (e.g., if they are familiar with the work of some faculty or have talked to them). These notes are entered into the evaluation system.

@@

Most of what is written in this document applies to both domestic\footnote{For simplicity domestic means you did your undergrad (or MS) at a US university. Different universities might have different definitions, e.g., permanent residents or US citizens.} and international students.  However, there are some differences and benefits that domestic students should be aware of and can leverage to improve their chances of admission.
\paragraph{Standing out \autoref{sec:improve-your-chance}} There are \emph{few} domestic applications compared to international ones, i.e., domestic students are the minority (\autoref{sec:urm}) in the CS PhD application pool. Many US universities want to increase the number of domestic students in their programs (and as mentioned later, there are specific fellowships and funding for domestic students).
That makes you stand out and can help your case.

@@

While both MS and PhD programs are graduate degrees, they are \emph{very different} in terms of objective, admission requirements, course requirements, duration, and funding. This section discusses the differences and provides guidance on applying to MS programs.

@@

In most cases MS CS programs are much less competitive than PhD programs, i.e., you're likely to get in if you can afford it. Many think of MS programs as a \emph{cash cow} because students are often not funded and have to pay tuition.

@@

\subsection{Funding}\label{sec:ms-funding}
Unlike PhD programs, which often have funding (\autoref{sec:funding}), MS students are typically not funded. The reason is because MS students do not focus on research and thus are not funded through RA and PhD students have priority for TA positions. 
\begin{itemize}
\item[\textbf{RA}] Profs. are not willing to take MS students as RAs because they are not around long enough to be productive. It can take a while for a student to get up to speed and start being productive, and by that time, they are about to graduate. Moreover, the goal of most MS students is to get a job, not to do research, so they are not as motivated to do research.

\item[\textbf{TA}] MS students are typically not given TA positions because PhDs are given priority. Many depts. do not have enough TAs for PhD students and so they cannot afford to give them to MS students.
\end{itemize}

@@

\item \textbf{Living Cost:} MS students are typically not funded, so living cost can be a big factor. You should consider the tuition and living expenses of the area. Note that sometimes living costs, e.g., renting, seem scary at first but students often find way to make it work, e.g., by sharing an apartment with other students.
